---
title: "FHLogReg2"
author: "Nionella Bt Stephen Sampil"
date: "2023-12-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

## Group Members

Nionella binti Stephen Sampil

Wan Nor Syafiqah binti Wan Salleh

Nur Hafizah binti Sukeri

Farah Munirah binti Mior Mazlan

Zahid bin Zulkifli

Ahmad Firdaus bin Mohamed

## Dataset

This dataset is publically available on the Kaggle website, it is an
ongoing cardiovascular study on residents of the town of Framingham,
Massachusetts. The dataset provides the patient's information. It
includes over 4000 records and 15 attributes.

Variables:

Each attribute is a potential risk factor. There are both demographic,
behavioral and medical risk factors.

**Demographic:**

-   Sex: male or female(Nominal)

-   Age: Age of the patient;(Continuous )

**Behavioral**

-   Current Smoker: whether or not the patient is a current smoker
    (Categorical/Nominal)

-   Cigs Per Day: the number of cigarettes that the person smoked on
    average in one day.(continuous)

**Medical( history)**

-   BP Meds: whether or not the patient was on blood pressure medication
    (Categorical/Nominal)

-   Prevalent Stroke: whether or not the patient had previously had a
    stroke (Categorical/Nominal)

-   Prevalent Hyp: whether or not the patient was hypertensive
    (Categorical/Nominal)

-   Diabetes: whether or not the patient had diabetes
    (Categorical/Nominal)

**Medical(current)**

-   Tot Chol: total cholesterol level (Continuous)

-   Sys BP: systolic blood pressure (Continuous)

-   Dia BP: diastolic blood pressure (Continuous)

-   BMI: Body Mass Index (Continuous)

-   Heart Rate: heart rate (Continuous)

-   Glucose: glucose level (Continuous)

    **Predict variable (desired target)**

-   10 year risk of coronary heart disease CHD (binary: "1", means
    "Yes", "0" means "No")

# Method

World Health Organization has estimated 12 million deaths occur
worldwide, every year due to Heart diseases. Half the deaths in the
United States and other developed countries are due to cardio vascular
diseases. The early prognosis of cardiovascular diseases can aid in
making decisions on lifestyle changes in high risk patients and in turn
reduce the complications. This research intends to pinpoint the most
relevant/risk factors of heart disease as well as predict the overall
risk using logistic regression.

### Prepare environment / Load libraries

```{r}
library(haven)
library(tidyverse)
library(gtsummary)
library(broom)
library(knitr)
library(tidyr)
library(dplyr)
library(corrplot)
library(dagitty)
```

### Read data

```{r}
Data1 <- read_sav("framinghamTrue.sav")
```

### View Data

```{r}
summary(Data1)
str(Data1)
glimpse(Data1)
```

### Transform data

```{r}
Data1 <- Data1 %>%
  mutate(across(where(is.labelled), as_factor)) %>%
  mutate(glucose = as.integer(glucose))
summary(Data1)
```

### Describe Data

```{r}
Data1 %>%
  tbl_summary(by = TenYearCHD, statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{n} / {N} ({p}%)"))
```

### Explore data

#### Plots

Histogram for numerical Barplots for categorical

Numerical: Age, CigPerday, totchol, SBP, DBP, BMI, HR, glucose
Categorical: Gender, Education, current smoker, BPmeds, PrevStroke,
PrevHPt, DMstat, TenYearCHD

```{r}
Data1 %>% 
  ggplot(aes(age)) + 
  geom_histogram() + 
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>%
  ggplot(aes(cigsPerDay)) +
  geom_histogram() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>%
  ggplot(aes(totChol)) +
  geom_histogram() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>%
  ggplot(aes(sysBP)) +
  geom_histogram() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>%
  ggplot(aes(diaBP)) +
  geom_histogram() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>%
  ggplot(aes(BMI)) +
  geom_histogram() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>%
  ggplot(aes(heartRate)) +
  geom_histogram() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>%
  ggplot(aes(glucose)) +
  geom_histogram() +
  facet_grid(. ~ TenYearCHD)
```

Barplots

```{r}
Data1 %>% 
  ggplot(aes(gender)) + 
  geom_bar() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>% 
  ggplot(aes(education)) +  
  geom_bar() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>% 
  ggplot(aes(currentSmoker)) +  
  geom_bar() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>% 
  ggplot(aes(BPMeds)) +  
  geom_bar() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>% 
  ggplot(aes(prevalentStroke)) +  
  geom_bar() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>% 
  ggplot(aes(prevalentHyp)) +  
  geom_bar() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>% 
  ggplot(aes(diabetes)) +  
  geom_bar() +
  facet_grid(. ~ TenYearCHD)
```

```{r}
Data1 %>% 
  ggplot(aes(TenYearCHD)) +  
  geom_bar()
```

#### Check Multicollinearity

```{r}
Data2 <- 
  Data1 %>% 
  select(where(is.numeric))
```

```{r}
cor.Data2 <-
  cor(Data2, use = "complete.obs", method = "pearson")
head(round(cor.Data2,2))
```

```{r}
corrplot(cor.Data2, type = 'upper', order = 'hclust')
```

From the correlation plot and correlation matrix, sbp and dbp is highly
correlated. Furthermore, the categorical variable of prevalent
hypertension already represent the hypertension status. Hence, sbp and
dbp are removed from model.

Similiarly, glucose is already representated by diabetes status.

#### Counfounder and mediator

#### Causal diagram

```{r}
myDag <- dagitty(' dag {
bb="0,0,1,1"
"Anti-Hypertensive Medication" [pos="0.602,0.881"]
"Smoking Status" [pos="0.743,0.775"]
"Total Cholesterol" [pos="0.302,0.122"]
Age [pos="0.541,0.072"]
BMI [pos="0.700,0.190"]
CHD [outcome,pos="0.817,0.482"]
Diabetes [exposure,pos="0.092,0.449"]
Gender [pos="0.185,0.754"]
Hypertension [pos="0.359,0.897"]
"Anti-Hypertensive Medication" -> CHD
"Anti-Hypertensive Medication" <-> Hypertension
"Smoking Status" -> CHD
"Smoking Status" -> Hypertension
"Total Cholesterol" -> CHD
"Total Cholesterol" -> Hypertension
Age -> CHD
Age -> Diabetes
Age -> Hypertension
BMI -> "Total Cholesterol"
BMI -> CHD
BMI -> Hypertension
BMI <-> Diabetes
Diabetes -> "Total Cholesterol"
Diabetes -> CHD
Diabetes -> Hypertension
Gender -> "Smoking Status"
Gender -> "Total Cholesterol"
Gender -> CHD
Gender -> Diabetes
Gender -> Hypertension
Hypertension -> CHD
}')
plot(myDag)
```

From the causal diagram, the exposure studied is diabetes
status(diabetes) and the outcome of interest is ten years risk of
developing CHD(TenYearCHD)

We identified that variables Age (age), BMI (BMI), Gender(Gender) as
confounder. Total cholesterol(totChol), hypertension(prevalentHyp) and
antihypertensive medication(BPMeds) as mediators from the causal
diagram.

# Results

## Estimation

## Simple Logistic Regression

Outcome: Ten Year Risk Developing CHD

Based on literature review and expert consultation, these are the
potential risk factors for developing Coronary heart disease in ten
years

Numerical: Age, total chol, BMI, CigsPerday, sbp, dbp, glucose, HR\
Categorical: HPT, DM, gender, BP meds, prev stroke, current Smoking
status, education

```{r}
#slr.all <- glm(TenYearCHD ~ 1, data = Data1, family = binomial(link = "logit"))
#add1(slr.all,scope = ~ age + totChol + BMI + cigsPerDay + sysBP + diaBP + heartRate + glucose + gender + education + prevalentHyp + prevalentStroke + currentSmoker + BPMeds + diabetes, test = "LRT")
```

#### ALL VARIABLES

```{r}
tbl_uvregression(Data1, method = glm, y = TenYearCHD, method.args = list(family = binomial), exponentiate = TRUE)
```

## Multiple Logistic Regression

### Model without interaction

Outcome: Ten Year risk of CHD

According to literature review, based on causal diagram and correlation
matrix, these are the potential risk factors for developing CHD:

Numerical IV : age + totChol + BMI + cigsPerDay\
Categorical IV: gender + prevalentHyp + BPMeds + diabetes

```{r}
mlr.main <- glm(TenYearCHD ~ gender + age + cigsPerDay + prevalentHyp + diabetes + totChol + BMI + BPMeds, family = binomial(link="logit"), data = Data1)
summary(mlr.main)
tidy(mlr.main)
tidy(mlr.main, exponentiate = TRUE)
```

```{r}
tbl_regression(mlr.main) %>%
  add_n(location = "level") %>%
  bold_labels()%>%
  italicize_levels()
```

```{r}
tbl_regression(mlr.main, exponentiate = TRUE) %>%
  add_n(location = "level") %>%
  bold_labels()%>%
  italicize_levels()
```

### Model with interaction

Outcome: Ten Year risk of CHD

Numerical IV : age + totChol + BMI + cigsPerDay\
Categorical IV: gender + prevalentHyp + BPMeds + diabetes

Interaction: Total Cholesterol with Hypertension status

```{r}
mlr.ia <- glm(TenYearCHD ~ gender + age + cigsPerDay + prevalentHyp + diabetes + totChol + BMI + BPMeds + totChol:prevalentHyp, family = binomial(link="logit"), data = Data1)
summary(mlr.ia)
tidy(mlr.ia)
tidy(mlr.ia, exponentiate = TRUE)
```

### Model Selection

Model without interaction vs model with interaction, need to compare
with anova test

```{r}
anova(mlr.main, mlr.ia, test = 'Chisq')
```

From the ANOVA The p-value (Pr(\>F)) is greater than the typical
significance level of 0.05. This suggests there is no difference between
the two models.

Hence, we choose multivariable model without interaction (mlr.main) as
our preliminary Final Model for Logistic Regression due to model with
the least variable is parsimonious model.

## Prediction

For prediction: 1. estimated log odds 2. probabilities 3. residuals 4.
hat values 5. Cooks distance 6. standardized residuals

Now, our preliminary final model is (mlr.main)

```{r}
prem.final.mod <- mlr.main
prem.final.mod
```

### Log Odds

```{r}
tidy(prem.final.mod, conf.int = TRUE)
```

### Predicted Odds

```{r}
tidy(prem.final.mod, exponentiate = TRUE, conf.int = TRUE)
```

### Predict Probability

Formula for odds = probability / 1 - probability Formula for probability
= Odds / 1 + Odds

```{r}
Prob.CHD <- augment(mlr.main, type.predict = 'response')  #To get probability, must include type.predict = 'response', if not, the fitted values will produce the log odds
Prob.CHD
```

```{r}
head(model.matrix(prem.final.mod))
```

```{r}
head(predict(prem.final.mod, type = 'link'))
```

```{r}
head(predict(prem.final.mod, type = 'response'))
```

Try to manually calculate:

Gender = 1 Age = 39 Cigs Perday = 0 HPT = 0 DM = 0 Totchol = 195 BMI =
26.97 BPmed = 0

Formula for log odds = B0 + B1X1 + B2X2 .......

Predicted log odds for first observation =

```{r}
(-7.1392+(0.4976*1)+(39*0.0719)+(0*0.0196)+(0*0.7249)+(195*0.0025)+(26.97*0.0152)+(0*0.3491))
```

Predicted odds:

Formula: exp(B0 + B1X1 + B2X2......)

```{r}
exp(-2.940056)
```

Manual calculation of predicted probability:

Formula for probability: Odds / 1 + Odds

```{r}
0.05286277/(1 + 0.05286277)
```

Another example: !!!! Try for 5th observation!!! Practice!!!

```{r}
augment(prem.final.mod, type.predict = 'response', type.residuals = 'pearson')
```

### Predict for new Data

```{r}
summary(model.matrix(prem.final.mod))
```

Make a predictions on new data

Example, what is the log odds, OR and probability of CHD for those aged
45 years old, tot chol 400, and with diabetes

```{r}
newdata1 <- expand.grid(age = seq(from = 32, to = 70, by = 10),
                       totChol = seq(from = 113, to = 400, by = 50),
                       diabetes = c('no dm', 'dm'),
                       gender = c('male', 'female'),
                       cigsPerDay = seq(from = 0, to = 70),
                       prevalentHyp = c('hpt', 'no hpt'),
                       BMI = seq(from = 15.54, to = 56.80, by=10),
                       BPMeds = c('not on hypertensive medication', 'on hypertensive medication'),
                       TenYearCHD = c('no chd', 'chd'))
newdata1
head(newdata1)
```

Example, what is the log odds, OR and probability of CHD for those aged
45 years old, tot chol 400, and with diabetes

```{r}
model.mlr.new <- glm(TenYearCHD ~ gender + age + cigsPerDay + prevalentHyp + diabetes + totChol + BMI + BPMeds, family = binomial(link="logit"), data = newdata1)
model.mlr.new
predict.odds <- augment(prem.final.mod, newdata = newdata1, type.predict = 'link')
head(predict.odds)
```

```{r}
predict.prob <- augment(prem.final.mod, newdata = newdata1, type.predict = 'response')
head(predict.prob)
```

```{r}
augment(prem.final.mod, newdata = newdata1, type.predict = 'response')
```

## Model Checking

### Check overall fitness:

1.  Accuracy
2.  Sensitivity
3.  SPecificity

### Linearity in logits

by using:

• area under the curve • Hosmer-Lemeshow test • modidied Hosmer-Lemeshow
test • Oseo Rojek test

#### Hosmer-Lemeshow Test

```{r}
# Assuming 'model' is your logistic regression model
library(ResourceSelection)
hoslem.test(Data1$TenYearCHD, fitted(prem.final.mod), g = 10)  # Adjust 'g' as needed

```

P value is \<0.001, hence model shows good fit

#### Area under the Curve (ROC)

```{r}
# Assuming 'model' is your logistic regression model
library(pROC)
roc_curve <- roc(TenYearCHD ~ predict(prem.final.mod, type = "response"), data = Data1)
auc(roc_curve)
plot(roc_curve)

```

Area under Receiver of Operating Characteristic Curve (AUC-ROC):

AUC-ROC is 73.1%, the model is considered moderate discriminating
effect.

### Diagnostic Plot

```{r}
plot(prem.final.mod)
```

From the Pearson residual plots, we can identify there are influential
outliers greater than +2

The studentized residual plots, we can also noted there are outliers
greater than +2, the leverage are higher than +2 suggests presence of
influential outliers

Hence, these outliers observations may affect considerably our logistic
regression model.

### Identify Influentials outliers

Need to identify influentials obs,


```{r}
# Assuming you have a linear regression model named 'your_model' and the data frame is named 'original_data'
# your_model <- lm(y ~ x1 + x2, data = original_data)

# Calculate influence measures
infl <- influence.measures(prem.final.mod)
infl.val <- data.frame(infl$infmat)

# Set a threshold for Cook's distance deviations
cutoff_value <- 4/3656

# Filter out influential observations
filtered_data <- Data1 |> filter(infl.val$cook.d <= cutoff_value)


# Fit the logistic regression model again using the filtered data
mlr.filtered.influential <- glm(TenYearCHD ~ gender + age + cigsPerDay + prevalentHyp + diabetes + totChol + BMI + BPMeds, family = binomial(link="logit"), data = filtered_data)

# Compare model summaries
summary(mlr.filtered.influential)
summary(prem.final.mod)
```

```{r}
tidy(prem.final.mod, conf.int = TRUE)
tidy(prem.final.mod, exponentiate = TRUE)
```

```{r}
tidy(mlr.filtered.influential, conf.int = TRUE)
tidy(mlr.filtered.influential, exponentiate = TRUE)
```
### Reperform Model fitness

#### Hosmer-Lemeshow Test

```{r}
# Assuming 'model' is your logistic regression model
library(ResourceSelection)
hoslem.test(filtered_data$TenYearCHD, fitted(mlr.filtered.influential), g = 10)  # Adjust 'g' as needed
```

P value <0.001, the filtered influential model maintains its model fitness.



#### Area under the Curve (ROC)

```{r}
# Assuming 'model' is your logistic regression model
library(pROC)
roc_curve <- roc(TenYearCHD ~ predict(mlr.filtered.influential, type = "response"), data = filtered_data)
auc(roc_curve)
plot(roc_curve)
```


Area under Receiver of Operating Characteristic Curve (AUC-ROC):

After model remodelling AUC-ROC is 82.26%, the model has improved to good discriminating effect.



### Reperform Diagnostic plots 

```{r}
plot(mlr.filtered.influential)
```

Most residuals appear to be falling within greater -/+ 2, lesser influentials observation. 

#### Compare between model without influential and preliminary final model(with influentials)

To identify the difference between B coefficients between these 2 models

```{r}
tidy(prem.final.mod, exponentiate = TRUE)
tidy(mlr.filtered.influential, exponentiate = TRUE)
```

Prelim Final Model --> B for age is 1.07, where as
Filtered Model --> B for age is 1.14
Due to the influentials, there are significant changes of B under diabetes status.

# Interpretation

Hence, the final model will be:

```{r}
Final.Model <- mlr.filtered.influential
```

```{r}
tbl_regression(Final.Model, exponentiate = TRUE) %>%
  bold_labels() %>%
  italicize_levels() %>%
  as_gt() %>%
  gt::tab_header(title = "Table 3. Multiple Logistic Regression (Final Model)")
```
Interpretation:

Gender, Age, Number of cigarette smoked per day, hypertension status have significant association to CHD.

1. Men has 2.36 time the odds compared to women to develop CHD (95% CI:1.75,3.19) when adjusted to age, cigperday, hypertension status @ being male has higher chance (36%) of developing CHD compared to female when adjusted to age, cigperday, hypertension status.
2. Every 1 year increase of age, has 1.14 time the odds of developing CHD (95% CI: 1.11,1.16) when adjusted to gender, cigperday, hypertension status.
3. Every 1 stick of cigarette per day, has 1.02 time the odds of developing CHD (95% CI 1.01,1.03) when adjusted to gender, age, hypertension status
4. Those who diagnosed with hypertension, has 2.12 time the odds of developing CHD (95% CI 1.60,2.80) when adjusted to gender, age, cigperday.
5. Diabetes status, total cholesterol level, BMI and on Hypertensive medications or not, are not statistically significant association to developing CHD after adjusted to all other variables. 






